---
title: "Credit Risk Analysis"
author: "Jagger Robles-Holmes"
date: "2025-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(tidyverse)    # Data manipulation & visualization
library(caret)        # Machine learning
library(randomForest) # Random Forest model
library(ggplot2)      # Plotting
library(ROCR)         # ROC curves
library(corrplot)     # Collinearity
```

```{r}
# Load data
credit_data <- read.csv("C:/Users/jagge/Downloads/credit_risk_dataset.csv/credit_risk_dataset.csv")
```

```{r}
# Check structure
glimpse(credit_data)

# Summary statistics
summary(credit_data)

# Check for missing values
colSums(is.na(credit_data))
```

```{r}
# Cap person_age at 80
credit_data <- credit_data %>%
  mutate(person_age = ifelse(person_age > 80, 80, person_age))
```

```{r}
# Impute loan_int_rate with median
credit_data$loan_int_rate[is.na(credit_data$loan_int_rate)] <- median(credit_data$loan_int_rate, na.rm = TRUE)

# Impute person_emp_length with 0 (assuming missing = unemployed)
credit_data$person_emp_length[is.na(credit_data$person_emp_length)] <- 0
```

```{r}
credit_data <- credit_data %>%
  mutate(
    person_home_ownership = as.factor(person_home_ownership),
    loan_intent = as.factor(loan_intent),
    loan_grade = as.factor(loan_grade),
    cb_person_default_on_file = as.factor(cb_person_default_on_file),
    loan_status = as.factor(loan_status)
  )
```

```{r}
# Age distribution
ggplot(credit_data, aes(x = person_age)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of Applicant Age", x = "Age", y = "Count")

# Income distribution (log scale for readability)
ggplot(credit_data, aes(x = person_income)) +
  geom_histogram(fill = "green", alpha = 0.7) +
  scale_x_log10() +  # Log-transform due to skewness
  labs(title = "Distribution of Income (Log Scale)", x = "Income", y = "Count")
```

```{r}
# Loan status (target variable)
ggplot(credit_data, aes(x = loan_status)) +
  geom_bar(fill = c("green", "red")) +
  labs(title = "Loan Default Status", x = "Status (0 = Paid, 1 = Default)", y = "Count")

# Loan intent
ggplot(credit_data, aes(x = loan_intent)) +
  geom_bar(fill = "orange") +
  coord_flip() +  # Horizontal bars for readability
  labs(title = "Purpose of Loan", x = "Intent", y = "Count")
```

```{r}
# Income vs. Default
ggplot(credit_data, aes(x = loan_status, y = person_income)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Income vs. Loan Default", x = "Default Status", y = "Income (Log)")

# Loan amount vs. Default
ggplot(credit_data, aes(x = loan_status, y = loan_amnt)) +
  geom_boxplot() +
  labs(title = "Loan Amount vs. Default", x = "Default Status", y = "Loan Amount")
```

```{r}
# Default rate by home ownership
credit_data %>%
  group_by(person_home_ownership, loan_status) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count)) %>%
  ggplot(aes(x = person_home_ownership, y = percent, fill = loan_status)) +
  geom_col() +
  labs(title = "Default Rate by Home Ownership", x = "Home Ownership", y = "Proportion")

# Default rate by loan intent
credit_data %>%
  group_by(loan_intent, loan_status) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count)) %>%
  ggplot(aes(x = loan_intent, y = percent, fill = loan_status)) +
  geom_col() +
  coord_flip() +
  labs(title = "Default Rate by Loan Purpose", x = "Purpose", y = "Proportion")
```

```{r}
# Select numeric variables only
numeric_data <- credit_data %>%
  select(person_age, person_income, person_emp_length, loan_amnt, loan_int_rate, loan_percent_income, cb_person_cred_hist_length)

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")  # Handles any remaining NAs

# Plot correlations
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black")
```

```{r}
# Separate test and train data
set.seed(123)
train_index <- createDataPartition(credit_data$loan_status, p = 0.75, list = FALSE)
train_data <- credit_data[train_index, ]
test_data <- credit_data[-train_index, ]
```

```{r}
# log transformation reduces skewness
train_data$log_age <- log(train_data$person_age + 1)  # +1 to avoid log(0)
```

```{r}
# Easier for higher incomes to pay higher loans
train_data$income_loan_ratio <- train_data$person_income / train_data$loan_amnt
```

```{r}
# Highlights riskier categories
train_data$high_risk_purpose <- ifelse(train_data$loan_intent %in% c("DEBTCONSOLIDATION", "MEDICAL", "HOMEMPROVEMENT"), 1, 0)
```

```{r}
# Converts text to binary columns 
dummies <- dummyVars(~ person_home_ownership + loan_intent + loan_grade, data = train_data)
train_encoded <- predict(dummies, newdata = train_data) %>% as.data.frame()
train_data <- cbind(train_data, train_encoded)
```

```{r}
# Logistic is good for binary classification
logit_model <- glm(
  loan_status ~ person_age + person_income + loan_amnt + loan_int_rate + 
                cb_person_cred_hist_length + high_risk_purpose + income_loan_ratio,
  data = train_data,
  family = "binomial"
)

# Summary to check coefficients
summary(logit_model)
```

```{r}
logit_model_simple <- glm(
  loan_status ~ person_income + loan_amnt + loan_int_rate + high_risk_purpose,
  family = "binomial",
  data = train_data
)

# Check new summary
summary(logit_model_simple)
```

```{r}
# Create the same high_risk_purpose feature in test_data
test_data$high_risk_purpose <- ifelse(
  test_data$loan_intent %in% c("DEBTCONSOLIDATION", "MEDICAL", "HOMEMPROVEMENT"), 
  1,  # High-risk
  0   # Low-risk
)

# Predict probabilities on test data
test_data$pred_prob <- predict(logit_model_simple, newdata = test_data, type = "response")
head(test_data$pred_prob)  # Show first few predicted probabilities
```

```{r}
# Convert probabilities to binary predictions (threshold = 0.5)
test_data$pred_status <- ifelse(test_data$pred_prob > 0.5, 1, 0)

# Generate confusion matrix
conf_matrix <- confusionMatrix(as.factor(test_data$pred_status), as.factor(test_data$loan_status))
print(conf_matrix)
```

```{r}
pred <- prediction(test_data$pred_prob, test_data$loan_status)
perf <- performance(pred, "tpr", "fpr")
plot(perf, colorize = TRUE, main = "ROC Curve")
abline(a = 0, b = 1, lty = 2)

# Calculate AUC
auc_log <- performance(pred, "auc")@y.values[[1]]
print(paste("AUC:", round(auc_log, 3)))
```

```{r}
set.seed(123)

# Trying random forest to handle imbalance between paid and default
class_weights <- ifelse(train_data$loan_status == 1, 5, 1)  # 5x weight for defaults

rf_model_balanced <- randomForest(
  loan_status ~ person_age + person_income + loan_amnt + loan_intent + 
                cb_person_default_on_file + income_loan_ratio,
  data = train_data,
  ntree = 200,
  classwt = c("0" = 1, "1" = 5),  # Explicit class weights
  importance = TRUE
)
```

```{r}
# Recreate income_loan_ratio in test_data
test_data$income_loan_ratio <- test_data$person_income / test_data$loan_amnt

# Predict probabilities
test_data$rf_prob <- predict(rf_model_balanced, newdata = test_data, type = "prob")[, 2]

# Convert to binary predictions (threshold = 0.5)
test_data$rf_pred <- ifelse(test_data$rf_prob > 0.5, 1, 0)
```

```{r}
conf_matrix_rf <- confusionMatrix(as.factor(test_data$rf_pred), 
                                as.factor(test_data$loan_status))
print(conf_matrix_rf)
```

```{r}
pred_rf <- prediction(test_data$rf_prob, test_data$loan_status)
perf_rf <- performance(pred_rf, "tpr", "fpr")

plot(perf_rf, colorize=TRUE, main="ROC Curve - Balanced Random Forest")
abline(a=0, b=1, lty=2)

# Calculate AUC
auc_rf <- performance(pred_rf, "auc")@y.values[[1]]
print(paste("AUC:", round(auc_rf, 3)))
```

This project aimed to predict credit risk by analyzing loan applicant data to determine whether borrowers would default. Starting with a dataset of 32,581 loans, I performed exploratory data analysis to identify key trends: defaults were more common among renters, high-risk loan purposes (medical, debt consolidation), and applicants with higher loan amounts relative to income.

I built and compared two supervised learning models in R:

  1. Logistic Regression: Provided interpretable coefficients but struggled with class imbalance, catching         only 34% of defaults.

  2. Balanced Random Forest: Improved default detection to 54.5% by weighting defaults more heavily, while         maintaining 83.5% accuracy and a stronger AUC.

The final model demonstrates that income, loan amount, and interest rate are critical predictors of default. While the model trades off some false positives for better default detection, this is often preferable in banking—where missing risky loans is costlier than over-caution. Future improvements could include fine-tuning the decision threshold or engineering additional features such as debt-to-income ratios.

This project highlights how machine learning can uncover actionable insights in financial risk, providing a foundation for real-world credit scoring systems. By methodically cleaning data, engineering features, and iterating on models, I transformed raw data into a predictive tool with measurable business value.



